\documentclass[11pt]{article}
\input{preamble}
  
%\newcommand{\ans}[1]{\emph{Solution: #1}}

\newcommand{\ans}[1]{}

\begin{document}

\title{CS 362, HW2}

\author {Prof. Jared Saia, University of New Mexico}

\date{\emph{Due: February 16th}}
\maketitle

If you have an older version of the book, check with a friend to make
sure you get the right numbers for exercises (the numbers are
different for the first and second editions)

\begin{enumerate}

\item 
Consider the recurrence $T (n) = 2T (n/3) + n^{2}$
\begin{enumerate}
\item Use the recursion tree method to get a tight upper bound
(i.e. big-O) on the solution to this recurrence
\item Now use annihilators (and a transformation) to get a tight upper
bound on the solution to this recurrence.  Show your work.  (Note that
your two bounds should match)
\end{enumerate}

\item 
Consider the recurrence $T (n) = 2T (n/2) + \log^{2} n$
\begin{enumerate}
\item Use the Master method to get a general solution to this
recurrence.
\item Now use annihilators (and a transformation) to get a tight upper
bound on the solution to this recurrence.  Show your work.  (Note that
your two bounds should match)
\end{enumerate}

\item Consider the following function:
\begin{verbatim}
int f (int n){
  if (n==0) return 3;
  else if (n==1) return 5;
  else{
    int val = 2*f (n-1);
    val = val - f (n-2);
    return val;
  }
}
\end{verbatim}

\begin{enumerate}
\item Write a recurrence relation for the \emph{value} returned by
$f$.  Solve the recurrence exactly.  (Don't forget to check it)
\item Write a recurrence relation for the \emph{running time} of $f$.
Get a tight upperbound (i.e. big-O) on the solution to this
recurrence. 
\end{enumerate}

\item Give the table for the optimal string alignment of the strings
``abbbaa'' and ``ababbab''.  Make sure you include arrows in your
table the same way we did in the table given in lecture, these arrows
will help you reconstruct the solution.  List all optimal alignments
of these two strings.

\item Find the optimal parenthesization for a matrix-chain product
whose sequence of dimensions is: $(2,3,2,2,1)$.  (Don't forget to
include the table used to compute your result)

\item Exercise 15.2-3

\item Exercise 15.2-5 (hint: use proof by induction)

\item Exercise 15.4-1

\item Exercise 15.4-5 (note: monotonically increasing means
non-decreasing, e.g. 1, 2, 2, 4, 5, 5, 7)

\item Extra Credit: Exercise 15.4-6

\item Extra Credit: Problem 15-2

\end{enumerate}

\end{document}
